\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-07 manuscript.Rmd

keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "1156"
bibliography      : "references.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional or demographic correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014]; as well as in males as compared to females [@miller2001; @papageorgiou2020; @lo2011; @shaqiri2018] (**Is there something for personality that we could use here instead of sex?**) - is another area of controversy.

@@

One key challenge hindering the further development of illusion research is the relative difficulty in adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. 
To address this issue, we first developed a parametric framework to manipulate visual illusions, which we implemented and made accessible in the open-source software *Pyllusion* [@makowski2021]. This software allows us to generate different types of classic visual illusions (e.g., Müller-Lyer, Ponzo, Delboeuf, Ebbinghaus) with a continuous and independent modulation of two parameters: *illusion strength* and *task difficulty* (see **Figure 1**).

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. **TODO: Insert a bit of literature about why and some findings justifying what we did**.

@@

Despite the widespread interest in finding associations between personality correlates and a general sensitivity to visual illusions, the supporting literature remains relatively mixed. Whereas some researchers report no relationship between inter-individual traits and individuals' susceptibility to illusions [@cretenoud2020a; @grzeczkowski2017], others have found pathological traits such as higher Aggression-Hostility and Narcissism, and lower Impulsive Sensation-Seeking to be significantly correlated to a greater resistance against such illusory effects [@zhang2017; @ohmann2016]. Furthermore, research on patients with schizophrenia have also generally demonstrated lower illusion sensitivity to be linked to symptoms of the disorder, such as greater hostility as well as social and emotional withdrawal [@pessoa2008].

@@

>>>>

keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "1156"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014; @pessoa2008], as well as in individuals with stronger traits of aggression and narcissism [@zhang2017; @konrath2009seeing] - is another area of controversy.

@@

One key challenge hindering the further development of illusion research is the relative difficulty in adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. To address this issue, we first developed a parametric framework to manipulate visual illusions, which we implemented and made accessible in the open-source software *Pyllusion* [@makowski2021]. This software allows us to generate different types of classic visual illusions (e.g., Müller-Lyer, Ponzo, Delboeuf, Ebbinghaus, ...) with a continuous and independent modulation of two parameters: *illusion strength* and *task difficulty* (see **Figure 1**).

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. Despite the abundant literature on visual illusions, relatively few studies investigate the ties with participants' characteristics. **TODO: HERE mention studies on sex or age**. Furthermore, most works assessing the links with personality focused on traits associated with psychopathology [such as impulsivity or sensation-seeking, @hlavata2018; @zhang2017; @lanyi2022can; @pessoa2008]. In this study, we instead propose to explore the relationship between illusion sensitivity, demographic variables and stable "general" personality traits.

@@

<!-- Context -->
We report significant links between inter-individual indices of illusion sensitivity and varialbes related to experimental context, demographic characteristics and personality. Firstly, screen size was found to have a significant negative relationship with the sensitivity to the two contrast-based illusions, namely the White and Contrast illusions. One possible explanation can be found in the mechanism by which visual systems filter through more low spatial frequencies when the size of the target object is small [@dixon2014]. As this filtering process excludes illumination information from visual processing, smaller screen sizes could yield artifactual changes in brightness perception, which in turn could attenuate the illusory effect of luminance-related illusions.

<!-- Demographic -->
**TODO: talk about absence of effect on sex + effect on age**

<!-- Normal personality -->
Our findings also suggested a positive relationship between illusion sensitivity and two "normal" personality traits, namely *Honesty-Humility* and *Agreeableness*. Although the past literature regarding the links between illusion sensitivity and Big 5 or HEXACO personality traits remain scarce, convergent evidence can be found in studies reporting a negative relationship between illusion sensitivity and hostility, aggression and narcissism [@zhang2017; @konrath2009seeing]. **TODO: do we have any leads of explanations?**

<!-- Pathological personality -->
The positive relationship between illusion sensitivity and "positive" personality traits is mirrored by a negative relationship with several pathological traits, including *Antagonism*, *Psychoticism*, *Disinhibition*, and *Negative Affect*. These results are in general consistent with the past findings and theories, suggesting that ...**TODO: do we have a theory?**

Our results, suggesting the existence of a link between illusion sensitivity and maladaptive personality traits in non-clinical populations, could be put in relation with clinical findings, which could be extreme cases where the relationship with illusion sensitivity is the most manifest. In line with our results on *Psychoticism* or *Disinhibition*, prior research has found greater illusion resistance in schizophrenia [**REF**], and in particular in association with schizotypal traits, such as cognitive disorganization [@cretenoud2019].

<!-- The results reported by past studies on the relationship between illusion sensitivity and personality remain fairly mixed [@cretenoud2020a; @grzeczkowski2017].  -->

<!-- Moreover, low aggression and high self-control have also been found to correlate strongly with greater resistance to illusions [@mcgurk]. -->

@@

<!-- Despite the widespread interest in finding associations between personality correlates and a general sensitivity to visual illusions, the supporting literature remains relatively mixed. Whereas some researchers report no relationship between inter-individual traits and individuals' susceptibility to illusions [@cretenoud2020a; @grzeczkowski2017], others have found pathological traits such as higher Aggression-Hostility and Narcissism, and lower Impulsive Sensation-Seeking to be significantly correlated to a greater resistance against such illusory effects [@zhang2017; @konrath2009seeing]. Furthermore, research on patients with schizophrenia have also generally demonstrated lower illusion sensitivity to be linked to symptoms of the disorder, such as greater hostility as well as social and emotional withdrawal [@pessoa2008]. -->
